### 2022-11-15 chapter1.tex
Changed minor awkward wording
Re-wrote section on connected components to include more discussion of structure of network

\subsection*{Input genomes and pre-processing}
All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database. The assemblies annotated by the NCBI eukaryotic genome annotation pipeline have passed several quality checks and all have supporting transcript evidence, so the annotations are generally highly complete~\cite{Thibaud-Nissen2013}. The \textit{D. miranda} annotation was excluded due to its unusual karyotype~\cite{Dobzhansky1935}. Other annotations were excluded after preliminary clustering showed a deficiency in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown). The \textit{D. melanogaster} annotation was downloaded from FlyBase~\cite{Gramates2022}. The complete list of genomes is given in Table~\ref{stable:genomes}. Many genes have transcripts that differ only in their UTRs, and as a result there are many duplicate protein sequences in the annotations. Though not strictly necessary, we removed the duplicates in our pipeline, which greatly reduced the computational burden of later steps.

@@

\subsection*{Clustering in orthologous groups}
% TODO: Add description of structure of connected components to motivate clustering algorithm
The best hits between sequences are naturally visualized as a graph where the sequences are nodes and best hits are edges between nodes (Fig.~\ref{fig:components}). Some sets of nodes form tight clusters whereas others are more loosely associated or even isolated. Automated genome annotation is not completely accurate, so some poorly supported hits are likely spurious, arising from homologous regions that were falsely annotated as genic in two closely related genomes. Others are possibly novel genes. Regardless of their origin, these hits are not strong candidates for comparative analyses since an orthology relationship is supported by relatively few genome pairs. In contrast, likely orthologs should consistently identify each other as best reciprocal hits across many genome pairs. Thus, orthologous groups can be operationally defined as self-consistent clusters in the hit graph. Since sequence divergence or assembly and annotation errors may prevent a best reciprocal hit between orthologous sequences across all genome pairs, the clustering method should require a high degree of self-consistency without demanding complete consensus.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{components/out/0022-002F.png}
\centering
\caption{\textbf{Selected connected components of hit graph.}
The best hits are shown as a graph where nodes are sequences and edges are best hits between sequences that are colored by their bit score. Two connected components, sets of nodes joined by a sequence of edges, are shown. The sequences in panels A are annotated as potassium channels whereas the sequences in panel B are uncharacterized. Some sets of nodes have a high density of edges, forming distinct clusters. Other nodes are only sparsely connected to their neighbors. Components can contain nodes of both types.}
\label{fig:components}
\end{figure}

@@

\subsection*{Grouping orthologous groups by gene}
As genes can have several annotated isoforms, each gene can be associated with several orthologous groups. However, the orthologous groups are not organized into gene-level units since they were clustered using sequence similarity only. A graph-based approach was therefore used to group orthologous groups with similar sets of parent genes. First, a gene overlap graph was constructed by defining an edge between orthologous groups if the intersection of their associated sets of parent genes is at least 50\% of the smaller of the two. Gene groups were then taken as the connected components of the resulting graph.

@@

Though the phylo-HMM removed phylogenetically incongruent ``insertions'', some sequences still contained extensive segments of uninterrupted gaps. These segments are easily identified in regions which are otherwise highly conserved, so they are also likely the result of incorrect or incomplete annotations. Unfortunately, they can also span more divergent regions, which complicates a simple rule- or threshold-based approach to identifying them. Thus, another phylo-HMM was trained to label each position in a sequence as generated by either a ``missing'' or ``not missing'' state. In this case, however, the aligned sequences were processed individually and not as aligned columns. As the previous phylo-HMM already ensured each column has sufficient support, these labels can instead be used to exclude sequences from downstream analyses depending on the amount of tolerated overlap with the regions of interest.

>>>>

\subsection*{Input genomes and pre-processing}
All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database. The assemblies annotated by the NCBI eukaryotic genome annotation pipeline have passed several quality checks and all have supporting transcript evidence, so the annotations are generally highly complete~\cite{Thibaud-Nissen2013}. The \textit{D. miranda} annotation was excluded due to its unusual karyotype~\cite{Dobzhansky1935}. Other annotations were excluded after preliminary clustering showed a deficiency in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown). The \textit{D. melanogaster} annotation was downloaded from FlyBase~\cite{Gramates2022}. In total, the input data consists of 33 genomes, which are listed in Table~\ref{stable:genomes}. Many genes have transcripts that differ only in their UTRs, and as a result there are many duplicate protein sequences in the annotations. Though not strictly necessary, we removed the duplicates in our pipeline, which greatly reduced the computational burden of later steps.

@@

\subsection*{Clustering in orthologous groups}
The best hits between sequences are naturally visualized as a graph where sequences are nodes and best hits are edges between nodes. Two connected components, sets of nodes joined by a sequence of edges, are shown in Figure \ref{fig:components}. The sequences in panel A all contain C2H2 zinc fingers whereas the sequences in panel B are members of the Par-1 family of serine/threonine protein kinases. In both components, some sets of nodes have a high density of edges, forming distinct clusters, whereas other nodes are only sparsely connected to their neighbors. To better understand the structure of these two components, we calculated the number of sequences, unique genes, and unique species in each. The first has 385, 346, and 33 sequences, genes, and species, respectively, and the second has 222, 33, and 33 sequences, genes, and species, respectively. We then plotted the relationship between the number sequences and unique genes across all components to see if this pattern holds true generally (Fig.~\ref{fig:components}C). Two distinct trendlines are apparent. The first increases linearly with the number of sequences with a slope of one, indicating each sequence is generally associated with a unique gene. The second is constant with an intercept of 33, indicating the number of unique genes quickly saturates at the total number of genomes. Thus, there are generally two classes of components. The first is composed of many distinct genes, whereas the second is composed of many different isoforms of a single group of genes.

The diffuse networks observed in the first component class are likely the result of a combination of factors, including rapid evolution, gene duplication, and annotation errors. Regardless of their origin, these hits are not strong candidates for comparative analyses since an orthology relationship is supported by relatively few genome pairs. Instead, likely orthologs should consistently identify each other as best reciprocal hits across many genome pairs. The same is true of the hits in the second component class. Although the genes as a unit form a single orthologous group, sequences with few hits are likely non-conserved or tissue-specific isoforms. Thus, orthologous groups can be operationally defined as self-consistent clusters in the hit graph. However, sequence divergence or assembly and annotation errors may prevent a best reciprocal hit between orthologous sequences across all genome pairs. In fact, although the most common number of reciprocal hits is 32, one fewer than the total number of genomes, many sequences have fewer (Fig.~\ref{fig:components}D). Thus, the clustering method should require a high degree of self-consistency without demanding complete consensus.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{components/out/0017-005F.png}
\centering
\caption{\textbf{Selected connected components of hit graph.}
Panels A and B are two distinct connected components of the graph formed by the best hits between sequences. Edges are colored by the value of their bit score.}
\label{fig:components}
\end{figure}

@@

\subsection*{Grouping orthologous groups by gene}
As genes can have several annotated isoforms, each gene can be associated with several orthologous groups. However, the orthologous groups are not organized into gene-level units since they were clustered using sequence similarity only. A graph-based approach was therefore used to group orthologous groups with similar sets of parent genes. First, a gene overlap graph was constructed by defining an edge between orthologous groups if the intersection of their associated sets of parent genes is at least 50\% of the smaller of the two. Gene groups were then taken as the connected components of the resulting graph.
% TODO: More description about results of this step

@@

Though the phylo-HMM removed phylogenetically incongruent ``insertions'', some sequences still contained extensive segments of uninterrupted gaps. These segments are easily identified in regions which are otherwise highly conserved, so they are also likely the result of incorrect or incomplete annotations. Unfortunately, they can also span more divergent regions, which complicates a simple rule- or threshold-based approach for identifying them. Thus, another phylo-HMM was trained to label each position in a sequence as generated by either a ``missing'' or ``not missing'' state. In this case, however, the aligned sequences were processed individually and not as aligned columns. As the previous phylo-HMM already ensured each column has sufficient support, these labels can instead be used to exclude sequences from downstream analyses depending on the amount of tolerated overlap with the regions of interest.
